---
title: "PBC"
author: "Stephen Shannon"
date: "June 28, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=F, results='hide', warning=FALSE}
library(dplyr)
library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
library(vtable)
library(ggplot2)
library(ggfortify)
```

```{r data, message=F, results='hide', warning=FALSE}
  #load data
data(pbc, package="survival")
```
Question 1. Assess the normality of each of the candidate variables and create a table showing the appropriate summary statistics ( e.g. mean +- sd or median and interquartile range)

```{r normality assumptions, message=F, results='hide', warning=FALSE}
  #check normality of the indepedent vars
par(mfrow=c(2,2))
qqnorm(pbcseq$bili, main = "Normal Q-Q Plot of Bilirubin");qqline(pbcseq$bili) #non-normal
qqnorm(pbcseq$albumin,main = "Normal Q-Q Plot of Albumin");qqline(pbcseq$albumin) #normal
qqnorm(pbcseq$alk.phos,main = "Normal Q-Q Plot of Alkaline Phosphate");qqline(pbcseq$alk.phos) #non-normal
qqnorm(pbcseq$ast,main = "Normal Q-Q Plot of Aspartate Aminotransferase");qqline(pbcseq$ast) #non-normal
qqnorm(pbcseq$platelet,main = "Normal Q-Q Plot of Platelets");qqline(pbcseq$platelet) #non-normal
qqnorm(pbcseq$protime,main = "Normal Q-Q Plot of Protime");qqline(pbcseq$protime) #non-normal
qqnorm(pbcseq$age,main = "Normal Q-Q Plot of Age");qqline(pbcseq$age) #normal
qqnorm(pbcseq$chol,main = "Normal Q-Q Plot of Cholestrol");qqline(pbcseq$chol) #non-normal
```
Figures 1 - 8: qqplots for each continous variable in the dataset, only albumin, and age are approximately normal.

Table 1. Summary statistics for each continous variable in the dataset
```{r sumtable}
  #table of summary statistics using vtable package for sumtable
sumtable(pbcseq, vars = c("bili", "chol", "albumin","alk.phos", "ast", "platelet", "protime", "age"), summ =  c('notNA(x)', 'mean(x)',   'sd(x)', 'min(x)', 'pctile(x)[25]', 'median(x)',  'pctile(x)[75]', 'max(x)'), out = "return")
```

```{r wrangling, message=F, results='hide', warning=FALSE}

  #glmnet requires that there are no NAs, and that the event var is only 0 or 1.
  #Since 1s are given to liver transplant cases, we must filter those who recieved
  #liver transplants and replace it with death, 2.

  #It may be easier to remove cholestrol from the analysis as many entries
  #are missing, and would remove all of these incomplete cases from
  #the analysis

pbcseq <- pbcseq %>% select(-chol)
pbcseq <- pbcseq %>% drop_na() %>% filter(status != 1) 
pbcseq["status"][pbcseq["status"] == 2] <- 1
```


```{r cox, message=F, warning=FALSE}
first <- with(pbcseq, c(TRUE, diff(id) !=0))
last <- c(first[-1], TRUE)

  #setup start, stop times and outcome for coxph
  #if first checkup, choose 0 days, otherwise choose the current day
time1 <- with(pbcseq, ifelse(first, 1, day))

  #if the last checkup, choose the follow up time, otherwise choose the previous check   
  #up time (since first checkup is not considered)
time2 <- with(pbcseq, ifelse(last, futime, day[-1]))

  #if last checkup, choose the current status, else choose censored as the outcome
event <- with(pbcseq, ifelse(last, status, 0))

  #basic model from the data source page

#m1 <- coxph(Surv(time1, time2, event) ~ age + sex + log(bili), pbcseq)
#summary(m1)

  #coxph model with every coeffecient
  #transforming some continous variables to the natural log
m2 <- coxph(Surv(time1, time2, event) ~ trt + age + sex + ascites + hepato + spiders + edema + stage + log(bili) + log(albumin) + log(alk.phos) + log(ast) + log(platelet) + log(protime),pbcseq)
summary(m2)

  #general survival curve
autoplot(surv_fit(m2, data=pbcseq))

  #identifies bili, albumin, edema stage, and age as statistically signifigant variables
```
Figure 9. General survival curve from the cox regression model

```{r cox plot 1, message=F, results='hide', warning=FALSE}
  #can add different strata to view difference between groups
  #making a cut to view range of groups
pbcseq$bili3 <- cut(pbcseq$bili, c(0,1,2.5,40))
strata_m2 <- coxph(Surv(time1, time2, event) ~ trt + age + sex + ascites + hepato + spiders + edema + stage + (bili) + log(albumin) + log(alk.phos) + log(ast) + strata(bili3) + log(platelet) + log(protime),pbcseq)
autoplot(survfit(strata_m2))
```
Figure 10. Survival curve with strata added for bilirubin

```{r cox plot 2, message=F, results='hide', warning=FALSE}

  #stratified survival curve for edema status
strata2_m2 <- coxph(Surv(time1, time2, event) ~ trt + age + sex + ascites + hepato + spiders + strata(edema) + stage + log(bili) + log(albumin) + log(alk.phos) + log(ast) + log(platelet) + log(protime),pbcseq)


autoplot(surv_fit(strata2_m2, data=pbcseq))
```
Figure 11. Survival curve with strata added for edema status

Question 2. Create a standard Cox regression model for survival using each of the variables as a predictor. Then use LASSO regression to identify a parsimonious set of variables predictive of survival. Propose what you believe to be the best model for the prediction of survival.
```{r glmnet_lasso, message=F, warning=FALSE}

  #lasso with cox regression, start, stop, status triplet used
  #cross validation method for lambda selection

y <- Surv(time1, time2, event)
x <- model.matrix(y ~ trt + sex + ascites + hepato + spiders + edema + stage + age + log(bili) + log(albumin) + log(alk.phos) + log(ast) + log(platelet) + log(protime), pbcseq)

  #glmnet model will not work unless low values of lambda are specified, otherwise
  #cv.glmnet will choose lambda > 10000 which imposes an absurdly strong penalty
  #function, leaving no params left in the model. Thus it is necessary to choose
  #lambda values manually
  #inspecting the algorithm with trace.it = 2, the models are indeed converging to a single
  #value but they are slightly off the target warm up number, triggering the 
  #cox.fit algorithm did not converge warning

m4 <- cv.glmnet(x,y, family="cox", standardize = TRUE, lambda = c(0.5, 0.1, 0.05, 0.01, 0.005, 0.004, 0.0035, 0.003, 0.0025, 0.001), alpha=1)
coef(m4)



  #from this model, the best predictors appear to be sex, ascites, edema, stage, 
  #age, log(bili), log(albumin), log(platelet), log(protime). Of these factors,
  #only ascites, edema, log(bili), log(albumin), and log(protime) have
  #covariate scores greater than abs(0.1), while the other variables are still
  #included at the optimal value of lambda, their score has little effect
  #on the hazard ratio for a patient. Age should still be included as it
  #despite being 0.02 as it is not log transformed and ranges from 26 to 78.
  #comparing the two models, the penalized model with LASSO regression has
  #4 variables with coeffecients larger than 0.1, which greatly reduces
  #the complexity of the model. The lower model complexity can reduce potential
  #overfitting present in the unpenalized cox regression model, 
```

```{r cv.glmnet lambda selection}
plot(m4)
```
Figure 12. Partial likelihood Deviance for different lambda selections

```{r glmnet survival curve, message=F, results='hide', warning=FALSE}
  #survival curve for a subject with covariates equal to the means of each variable.
  #glmnet is not able to produce confidence intervals
plot(survival::survfit(m4, s = "lambda.min", x = x, y = y))
```
Figure 13. Survival curve generated from the L1 regularized glmnet model.

Question 3. Employ a valid way to assess whether any variables may be non-linearly associated with the time to event outcome. 

```{r nonlinear_associations, message=F, results='hide', warning=FALSE}
ggcoxfunctional(Surv(time1, time2, event) ~ bili + log(bili) + sqrt(bili), data = pbcseq) #close to linear, seems to be more linear with the log transformation or the square root transformation.
ggcoxfunctional(Surv(time1, time2, event) ~ albumin + log(albumin) + sqrt(albumin), data = pbcseq) #linear or close to linear, log form does not seem to improve the linearity
ggcoxfunctional(Surv(time1, time2, event) ~ alk.phos + log(alk.phos) + sqrt(alk.phos), data = pbcseq) #not linear, log transformation improves linearity graetly
ggcoxfunctional(Surv(time1, time2, event) ~ ast + log(ast) + sqrt(ast), data = pbcseq) #not linear, log transformation does not improve linearity, square root may be most linear
ggcoxfunctional(Surv(time1, time2, event) ~ platelet + log(platelet) + sqrt(platelet), data = pbcseq) #Close to linear, log transformation does not improve linearity much
ggcoxfunctional(Surv(time1, time2, event) ~ protime + log(protime) + sqrt(protime), data = pbcseq) #not linear, log transformation marginally helps, square root function not much different, from log, appears to have residual outliers that may be influencing the curve
ggcoxfunctional(Surv(time1, time2, event) ~ age + log(age) + sqrt(age), data = pbcseq) #linear or close to linear, virtually no change in curve with the transformations

```
Figure 14 - 20. Plots of Martingale residuals against the continous covariates. The closer the curve is to a linear form (consistent slope, generally monotonic increase or decrease), the closer the functional form is to showing a linear association between the covariate and the hazards ratio.

Question 4. Create a figure that you feel adequately displays the results of your analysis that you have found

```{r important visuals 1, message=F, warning=FALSE}

  #survival curve for two subjects close to the same age, one with edema and ascites and high     
  #bilirubin, the other with low bilirubin, no edema and no ascites, the patient with 
  #presence of edema and ascites, high bilirubin creates very strong hazard ratio

x[2:3,]
plot(survival::survfit(m4, s = "lambda.min", x = x, y = y, newx = x[2:3,]))
```
Figure 21. Survival curve of two patients described above, the patient with edema, ascites, and high bilirubin has a much higher hazard ratio than the patient with no edema, ascites and low bilirubin.
```{r important visuals 2, message=F, results='hide', warning=FALSE}
strata_m2 <- coxph(Surv(time1, time2, event) ~ strata(trt) + age + sex + ascites + hepato + spiders + edema + stage + log(bili) + log(albumin) + log(alk.phos) + log(ast) + log(platelet) + log(protime),pbcseq)

  #looks as if the treatment makes little difference compared
  #to no treatment

autoplot(survfit(strata_m2))

```
Figure 22. Two survival curves based on the strata of the treatment levels. Treatment makes little to no difference compared to no treatment.
